# Reproducible Research: Peer Assessment 1

****
## Loading and preprocessing the data
```{r loaddata,echo=TRUE}
data <- read.csv("activity.csv")
data$date <- as.Date(data$date) ## convert character *date* to *date object*
str(data)
summary(data)
```
****
## What is mean total number of steps taken per day?
```{r meantotalsteps,echo=TRUE}
### tapply used to return a vector with the sum of steps for each day
### (the dimension names are the dates)
totalStepsByDay <- tapply(data$steps,data$date,sum)
totalStepsByDayMean <- mean(totalStepsByDay,na.rm=TRUE)
totalStepsByDayMedian <- median(totalStepsByDay,na.rm=TRUE)
hist(totalStepsByDay,
     main="Total Steps By Day",
     xlab="total steps by day")
abline(v=totalStepsByDayMean,col="blue",lwd=2,lty="dashed")
abline(v=totalStepsByDayMedian,col="yellow",lwd=2,lty="dashed")
```

### Total steps mean  : 
```{r,echo=TRUE} 
mean(totalStepsByDay,na.rm=TRUE)
```
### Total steps median: 
```{r,echo=TRUE} 
median(totalStepsByDay,na.rm=TRUE)
```


****
## What is the average daily activity pattern?
### 1. Time series plot of average daily activity pattern
```{r averageactivity1, echo=TRUE}
## tapply used to return a vector with the mean of steps for each interval,
## accross all dates
## (the dimension names are the intervals)
avgStepsByInterval <- tapply(data$steps,data$interval,mean,na.rm=TRUE)
plot(names(avgStepsByInterval),avgStepsByInterval,type="l",
     main="Average Daily Activity Pattern",
     ylab="Average Number Of Steps",
     xlab="Intervals")
```
### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r averageactivity2, echo=TRUE}
## the function "which.max" returns the index in the vector where the
## maximum number of averaged steps are found in the vector, and
## the "names" function return the named interval.
```
### Interval with maximum average of steps: 
```{r,echo=TRUE} 
names(which.max(avgStepsByInterval))
```


****
## Imputing missing values
### 1. Total number of missing values in dataset
Although there are only missing values in the "steps" column, I used the 
*complete.cases* function because they required **missing values in dataset** and not in a specific column
```{r imputingmissing1}
data2 <- data ## create a new data frame to work the missing values
indexMissingValuesInData <- complete.cases(data) ## index the complete cases
## used the function table to calculate missing and non missing values, 
## but printed only the number for missing as required
```
### Total of rows with missing values: 
```{r,echo=TRUE} 
table(indexMissingValuesInData)["FALSE"]
```

### 2. Filling in all of the missing values in the dataset
The strategy used to fill the missing values is substitute de NAs with then 
*mean* of each interval accross all days, reusing the vector 
*avgStepsByInterval* created in a previous section of the assignment.
The code for this strategy is in the next section.
### 3. Create a new dataset ...
```{r imputingmissing2}
## print summary of original dataset
summary(data)
## create new column in data frame to indicate where information is missing or 
## not. "TRUE" indicates a missing value, "FALSE" indicates the opposite
data2$missing <- !indexMissingValuesInData ## "not" to preserve meaning
## below is a for loop to fill the missing values with the mean of each
## interval accross all days
for (i in 1:nrow(data2)) {
        if (data2[i,"missing"]) {
                data2[i,"steps"] <- 
                        avgStepsByInterval[as.character(data2[i,"interval"])]
        }
}
## print summary of new dataset with missing values filled
summary(data2)
```
### 4. Make a histogram of the total number of steps taken each day after imputation of missing values
```{r meantotalsteps2,echo=TRUE}
### tapply used to return a vector with the sum of steps for each day
### (the dimension names are the dates)
totalStepsByDayData2 <- tapply(data2$steps,data2$date,sum)
hist(totalStepsByDayData2,
     main="Total Steps By Day (imputing NAs)",
     xlab="total steps by day")
```
### Total steps mean (imputing NAs):
```{r,echo=TRUE}
mean(totalStepsByDayData2)
```
### Total steps median (imputing NAs):
```{r, echo=TRUE}
median(totalStepsByDayData2)
```
### Difference and impact
The *mean* and *median* values differ from the ones obtained in the first part of the assignment, but only slightly. The real impact was in the frequency of steps in the histogram.

****
## Are there differences in activity patterns between weekdays and weekends?
### 1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r,echo=TRUE}
data3 <- data
week <- c("weekend",rep("weekday",5),"weekend") ## vector to provide factors
## define function to calculate weekday/weekend 
weekfun <- function(x) { 
        ## add 1 to wday (0-6 range) to make it compatible with vector range (1-7)
        week[as.POSIXlt(x)$wday+1] 
} 
## use *sapply* to apply previous function to all dates 
data3$week <- sapply(data3$date,weekfun,USE.NAMES=FALSE)
data3$week <- factor(data3$week) ## transform into factors
str(data3) ## show that the new data frame has a factor weekday|weekend
## create data frame to hold average of steps by interval by week
data4 <- aggregate(steps ~ week + interval,data=data3,FUN=mean)
```
### 2. Panel plot comparing weekends and weekdays activity pattern
```{r,echo=TRUE}
library(lattice)
xyplot(steps~interval|week,data4,
       type="l",
       layout=c(1,2),
       main="Weekends X Weekdays Activity Pattern",
       ylab="Average number of steps",
       xlab="Interval")
```
